---
title: "comparison_between_Th1_Th2_with_naive"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# read the data
```{r}
library(ggplot2)
file_list <- list.files(pattern = ".txt")
data_set <- lapply(file_list, function(x){read.delim(x,header = F,as.is = T,stringsAsFactors = F)})
info <- data_set[[1]]
#unlist(data_set)
data_fr <- as.data.frame(data_set[3:14])
data_fr <- data_fr[,grep("V2", colnames(data_fr))]
rownames(data_fr) <- data_set[3][[1]][,1]
colnames(data_fr) <- info[,2]
Th1 <- data_fr[ ,grep("Th1-",colnames(data_fr),fixed = T)]
Th2 <- data_fr[ ,grep("Th2-",colnames(data_fr),fixed = T)]
Th0 <- data_fr[ ,grep("Naive-",colnames(data_fr),fixed = T)]
Th17 <- data_fr[,grep("Th17",colnames(data_fr),fixed = T)]
iTreg <- data_fr[,grep("iTreg",colnames(data_fr),fixed = T)]
nTreg <- data_fr[,grep("nTreg",colnames(data_fr),fixed = T)]
df <- as.data.frame(cbind(Th0,Th1,Th2,Th17,iTreg,nTreg))
expression_level <- log2(df)
```
#PCA analysis for all T celltypes
```{r}
### run a PCA analysis for all T celltypes
PCA_data<- as.data.frame(t(expression_level))
PCA_data$group <- c("Naive", "Naive","Th1", "Th1","Th2", "Th2","Th17","Th17", "iTreg","iTreg", "nTreg","nTreg")
PCA_plot<- prcomp(PCA_data[,1:45101])
ggplot(data = NULL, aes(PCA_plot$x[,1],PCA_plot$x[,2],color=PCA_data$group) ) + geom_point() + xlab("First component") + ylab("Second component")+labs(color="celltype")
PoV <- PCA_plot$sdev^2/sum(PCA_plot$sdev^2)*100
barplot(PoV, xlab= "Dimensions", ylab="Proportion of explained variance (%)")
### genes with the top500 highest variance
PCA_var <- apply(PCA_data[,1:45101],2,var)
PCA_var <- order(PCA_var,decreasing = T)
PCA_var <- PCA_var[1:500]
PCA_var_data <- PCA_data[PCA_var]
PCA_var_plot <- prcomp(PCA_var_data)
ggplot(data = NULL, aes(PCA_var_plot$x[,1],PCA_var_plot$x[,2],color=PCA_data$group) ) + geom_point() + xlab("First component") + ylab("Second component")+labs(color="celltype")
PoV <- PCA_var_plot$sdev^2/sum(PCA_var_plot$sdev^2)*100
barplot(PoV, xlab= "Dimensions", ylab="Proportion of explained variance (%)")
```


# start eliminating T celltypes, no longer taking into the consideration of nTreg and iTreg
```{r}
### run a PCA analysis eliminating the influence of iTreg and nTreg

PCA_data<- as.data.frame(t(expression_level))
PCA_data$group <- c("Naive","Naive", "Th1","Th1", "Th2", "Th2","Th17","Th17", "iTreg", "iTreg", "nTreg","nTreg")
PCA_data <- PCA_data[1:8,]
PCA_plot<- prcomp(PCA_data[,1:45101])
ggplot(data = NULL, aes(PCA_plot$x[,1],PCA_plot$x[,2],color=PCA_data$group) ) + geom_point() + xlab("First component") + ylab("Second component")+labs(color="celltype")
PoV <- PCA_plot$sdev^2/sum(PCA_plot$sdev^2)*100
barplot(PoV, xlab= "Dimensions", ylab="Proportion of explained variance (%)")
### genes with the top500 highest variance
PCA_var <- apply(PCA_data[,1:45101],2,var)
PCA_var <- order(PCA_var,decreasing = T)
PCA_var <- PCA_var[1:500]
PCA_var_data <- PCA_data[PCA_var]
PCA_var_plot <- prcomp(PCA_var_data)
ggplot(data = NULL, aes(PCA_var_plot$x[,1],PCA_var_plot$x[,2],color=PCA_data$group) ) + geom_point() + xlab("First component") + ylab("Second component")+labs(color="celltype")
PoV <- PCA_var_plot$sdev^2/sum(PCA_var_plot$sdev^2)*100
barplot(PoV, xlab= "Dimensions", ylab="Proportion of explained variance (%)")

```
#Get the mean of different celltypes, not log2 transformed
```{r}
Th1 $ mean <- rowMeans(Th1)
Th2 $ mean <- rowMeans(Th2)
Th0 $ mean <- rowMeans(Th0)
Th17 $ mean <- rowMeans(Th17)
iTreg $ mean <- rowMeans(iTreg)
nTreg $ mean <- rowMeans(nTreg)
exp_Th0 <- expression_level[,1:2]
exp_Th0 $ mean <- rowMeans(exp_Th0)
exp_Th1 <- expression_level[,3:4]
exp_Th1 $ mean <- rowMeans(exp_Th1)
exp_Th2 <- expression_level[,5:6]
exp_Th2 $ mean <- rowMeans(exp_Th2)
exp_Th17 <- expression_level[,7:8]
exp_Th17 $ mean <- rowMeans(exp_Th17)
exp_iTreg <- expression_level[,9:10]
exp_iTreg $ mean <- rowMeans(exp_iTreg)
exp_nTreg <- expression_level[,11:12]
exp_nTreg $ mean <- rowMeans(exp_nTreg)
## generate the whole list for comparison
library(AnnotationDbi)
library(mouse4302.db)
whole_Th1 <- cbind(exp_Th1, exp_Th0)
whole_symbol<-mapIds(mouse4302.db,keys=as.character(row.names(whole_Th1)),column = "SYMBOL",keytype = "PROBEID" ) 
whole_Th1$symbol <- whole_symbol
whole_Th1<-whole_Th1[,c(7,1:6)]
colnames(whole_Th1) <- c("symbol","Th1_sample1","Th1_sample2","Th1_mean","Th0_sample1","Th0_sample2","Th0_mean")
whole_Th1$logFC <- whole_Th1$Th1_mean - whole_Th1$Th0_mean
whole_Th1$direction <- T
whole_Th1$direction[whole_Th1$logFC<0 ] <- F

whole_Th2 <- cbind(exp_Th2, exp_Th0)
whole_Th2$symbol <- whole_symbol
whole_Th2<-whole_Th2[,c(7,1:6)]
colnames(whole_Th2) <- c("symbol","Th2_sample1","Th2_sample2","Th2_mean","Th0_sample1","Th0_sample2","Th0_mean")
whole_Th2$logFC <- whole_Th2$Th2_mean - whole_Th2$Th0_mean
whole_Th2$direction <- T
whole_Th2$direction[whole_Th2$logFC<0 ] <- F

whole_Th17 <- cbind(exp_Th17, exp_Th0)
whole_Th17$symbol <- whole_symbol
whole_Th17<-whole_Th17[,c(7,1:6)]
colnames(whole_Th17) <- c("symbol","Th17_sample1","Th17_sample2","Th17_mean","Th0_sample1","Th0_sample2","Th0_mean")
whole_Th17$logFC <- whole_Th17$Th17_mean - whole_Th17$Th0_mean
whole_Th17$direction <- T
whole_Th17$direction[whole_Th17$logFC<0 ] <- F

write.table(whole_Th1,file = "~/R/Project_3_Lung Fibrosis/comparison_with_TH1_TH2/GSE14308_TH2_VS_TH1_UP_for_TH/GSE14308_family.xml/overall_expression_of_Th1.txt",quote = F)
write.table(whole_Th2,file = "~/R/Project_3_Lung Fibrosis/comparison_with_TH1_TH2/GSE14308_TH2_VS_TH1_UP_for_TH/GSE14308_family.xml/overall_expression_of_Th2.txt",quote = F)
write.table(whole_Th17,file = "~/R/Project_3_Lung Fibrosis/comparison_with_TH1_TH2/GSE14308_TH2_VS_TH1_UP_for_TH/GSE14308_family.xml/overall_expression_of_Th17.txt",quote = F)
```
# Try to rule out the influence of iTreg and nTreg, and compare the three subtypes with the Th0 to see the awareness behavior of different immune T cells
```{r}
comp_immuneawareness <- expression_level[,1:8]
library(limma)
design <- model.matrix(~ 0 + factor(c(1,1,2,2,3,3,4,4)))
colnames(design) <- c( "Naive","Th1","Th2","Th17")
fit <- lmFit(comp_immuneawareness ,design =design)
contrast_mat <-makeContrasts(Th1-Naive,Th2-Naive,Th17-Naive,levels = design)
fit2 <- contrasts.fit(fit, contrast_mat)
fit <-eBayes(fit2, 0.01)
p.adj<-p.adjust(fit$p.value,method = "fdr")
p.adj<-matrix(as.numeric(p.adj),ncol=3,nrow=45101)
library(AnnotationDbi)
library(mouse4302.db)
gen_symbol<-mapIds(mouse4302.db,keys =as.character(row.names(data_fr)), column = "SYMBOL",keytype = "PROBEID" ) 
T_awareness<-as.data.frame(cbind(gen_symbol,fit$p.value,p.adj))
names(T_awareness)<-c("Gene_symbol","pvalue_Th1","pvalue_Th2","pvalue_Th17","padj_Th1","padj_Th2","padj_Th17")
Th1_list <-T_awareness[as.numeric(as.character(T_awareness$padj_Th1))<= 0.05,]
Th1_list <- Th1_list[,c(1,2,5)]
Th1_list $ Th0 <- exp_Th0$mean[rownames(exp_Th0) %in% rownames(Th1_list)]
Th1_list $ Th1 <- exp_Th1$mean[rownames(exp_Th1) %in% rownames(Th1_list)]
Th1_list $ logFC <- Th1_list$Th1-Th1_list$Th0
Th1_list$direction <- T
Th1_list$direction[Th1_list $logFC <0] <- F
Th2_list <-T_awareness[as.numeric(as.character(T_awareness$padj_Th2))<= 0.05,]
Th2_list <- Th2_list[,c(1,3,6)]
Th2_list $ Th0 <- exp_Th0$mean[rownames(exp_Th0) %in% rownames(Th2_list)]
Th2_list $ Th2 <- exp_Th2$mean[rownames(exp_Th2) %in% rownames(Th2_list)]
Th2_list $ logFC <- Th2_list$Th2-Th2_list$Th0
Th2_list$direction <- T
Th2_list$direction[Th2_list $logFC <0] <- F
Th17_list <-T_awareness[as.numeric(as.character(T_awareness$padj_Th17))<= 0.05,]
Th17_list <- Th17_list[,c(1,4,7)]
Th17_list $ Th0 <- exp_Th0$mean[rownames(exp_Th0) %in% rownames(Th17_list)]
Th17_list $ Th17 <- exp_Th17$mean[rownames(exp_Th17) %in% rownames(Th17_list)]
Th17_list $ logFC <- Th17_list$Th17-Th17_list$Th0
Th17_list$direction <- T
Th17_list$direction[Th17_list $logFC <0] <- F
write.table(Th1_list,file = "~/R/Project_3_Lung Fibrosis/comparison_with_TH1_TH2/GSE14308_TH2_VS_TH1_UP_for_TH/GSE14308_family.xml/Th0 to Th1.txt",quote = F)
write.table(Th2_list,file = "~/R/Project_3_Lung Fibrosis/comparison_with_TH1_TH2/GSE14308_TH2_VS_TH1_UP_for_TH/GSE14308_family.xml/Th0 to Th2.txt",quote = F)
write.table(Th17_list,file = "~/R/Project_3_Lung Fibrosis/comparison_with_TH1_TH2/GSE14308_TH2_VS_TH1_UP_for_TH/GSE14308_family.xml/Th0 to Th17.txt",quote = F)
ggplot(mapping=aes(whole_Th2$logFC,-log10(as.numeric(as.character(p.adj[,2])))))+geom_point(size= 0.002,alpha=0.5)+ labs(x="log 2 fold change", y="-log10 p-value", title="Volcano Plot for Th2 cell differentiation")
ggplot(mapping=aes(whole_Th1$logFC,-log10(as.numeric(as.character(p.adj[,1])))))+geom_point(size= 0.002,alpha=0.5)+ labs(x="log 2 fold change", y="-log10 p-value", title="Volcano Plot for Th1 cell differentiation")
ggplot(mapping=aes(whole_Th17$logFC,-log10(as.numeric(as.character(p.adj[,3])))))+geom_point(size= 0.002,alpha=0.5)+ labs(x="log 2 fold change", y="-log10 p-value", title="Volcano Plot for Th17 cell differentiation")
```
# Another approach to the comparison, try to see whether only thinking about the Th1 Th2 and TH17 would make the seperation more meaningful
```{r}
# PCA analysis of only three different kinds of activated T helper cells
PCA_data<- as.data.frame(t(expression_level))
PCA_data$group <- c("Naive","Naive", "Th1","Th1", "Th2", "Th2","Th17","Th17", "iTreg", "iTreg", "nTreg","nTreg")
PCA_data <- PCA_data[3:8,]
PCA_plot<- prcomp(PCA_data[,1:45101])
ggplot(data = NULL, aes(PCA_plot$x[,1],PCA_plot$x[,2],color=PCA_data$group) ) + geom_point() + xlab("First component") + ylab("Second component")+labs(color="celltype")
PoV <- PCA_plot$sdev^2/sum(PCA_plot$sdev^2)*100
barplot(PoV, xlab= "Dimensions", ylab="Proportion of explained variance (%)") 
### genes with the top500 highest variance
PCA_var <- apply(PCA_data[,1:45101],2,var)
PCA_var <- order(PCA_var,decreasing = T)
PCA_var <- PCA_var[1:500]
PCA_var_data <- PCA_data[PCA_var]
PCA_var_plot <- prcomp(PCA_var_data)
ggplot(data = NULL, aes(PCA_var_plot$x[,1],PCA_var_plot$x[,2],color=PCA_data$group) ) + geom_point() + xlab("First component") + ylab("Second component")+labs(color="celltype")
PoV <- PCA_var_plot$sdev^2/sum(PCA_var_plot$sdev^2)*100
barplot(PoV, xlab= "Dimensions", ylab="Proportion of explained variance (%)") 
```

# When only having Th1 and Th2, because those are most specific for the research of lung fibrosis
```{r}
PCA_data<- as.data.frame(t(expression_level))
PCA_data$group <- c("Naive","Naive", "Th1","Th1", "Th2", "Th2","Th17","Th17", "iTreg", "iTreg", "nTreg","nTreg")
PCA_data <- PCA_data[3:6,]
PCA_plot<- prcomp(PCA_data[,1:45101])
ggplot(data = NULL, aes(PCA_plot$x[,1],PCA_plot$x[,2],color=PCA_data$group) ) + geom_point() + xlab("First component") + ylab("Second component")+labs(color="celltype")
PoV <- PCA_plot$sdev^2/sum(PCA_plot$sdev^2)*100
barplot(PoV, xlab= "Dimensions", ylab="Proportion of explained variance (%)") 
### genes with the top500 highest variance
PCA_var <- apply(PCA_data[,1:45101],2,var)
PCA_var <- order(PCA_var,decreasing = T)
PCA_var <- PCA_var[1:500]
PCA_var_data <- PCA_data[PCA_var]
PCA_var_plot <- prcomp(PCA_var_data)
ggplot(data = NULL, aes(PCA_var_plot$x[,1],PCA_var_plot$x[,2],color=PCA_data$group) ) + geom_point() + xlab("First component") + ylab("Second component")+labs(color="celltype")
PoV <- PCA_var_plot$sdev^2/sum(PCA_var_plot$sdev^2)*100
barplot(PoV, xlab= "Dimensions", ylab="Proportion of explained variance (%)") 
## When only having Th1 and Th2, the differences among samples of the same subtype would even influence the result of the comparison
```
# After analysis, it would be recommended that Th1, Th2 and Th17 as a group for the comparison between different activated T helper cells, generating an ANOVA similar result list
```{r}
diff_activ_Th <- expression_level[,3:8]
library(limma)
design <- model.matrix(~ 0 + factor(c(1,1,2,2,3,3)))
colnames(design) <- c( "Th1","Th2","Th17")
fit <- lmFit(diff_activ_Th ,design =design)
contrast_mat <-makeContrasts(Th17-Th1,Th2-Th1,Th17-Th2,levels = design)
fit2 <- contrasts.fit(fit, contrast_mat)
fit <-eBayes(fit2, 0.01,trend = T)
#fit <-eBayes(fit2, 0.01)
p.adj<-p.adjust(fit$p.value,method = "fdr")
p.adj<-matrix(as.numeric(p.adj),ncol=3,nrow=45101)
library(AnnotationDbi)
library(mouse4302.db)
gen_symbol<-mapIds(mouse4302.db,keys =as.character(row.names(data_fr)), column = "SYMBOL",keytype = "PROBEID" ) 
T_diff_act<-as.data.frame(cbind(gen_symbol,fit$p.value,p.adj))
names(T_diff_act)<-c("Gene_symbol","pvalue_Th17_Th1","pvalue_Th2_Th1","pvalue_Th17_Th2","padj_Th17_Th1","padj_Th2_Th1","padj_Th17_Th2")
T_1721_sig <- T_diff_act[as.numeric(as.character(T_diff_act$padj_Th2_Th1))<=0.05,]
write.table(T_1721_sig,file = "~/R/Project_3_Lung Fibrosis/comparison_with_TH1_TH2/GSE14308_TH2_VS_TH1_UP_for_TH/GSE14308_family.xml/List of Th1_Th2 when compared with Th17.txt",quote = F)

```
# Only using Th1 and Th2, expected is a shorter list of DE
```{r}
diff_activ_Th <- expression_level[,3:6]
library(limma)
design <- model.matrix(~ 0 + factor(c(1,1,2,2)))
colnames(design) <- c( "Th1","Th2")
fit <- lmFit(diff_activ_Th ,design =design)
contrast_mat <-makeContrasts(Th2-Th1,levels = design)
fit2 <- contrasts.fit(fit, contrast_mat)
fit <-eBayes(fit2, 0.01,trend = T)
p.adj<-p.adjust(fit$p.value,method = "fdr")
p.adj<-matrix(as.numeric(p.adj),ncol=1,nrow=45101)
library(AnnotationDbi)
library(mouse4302.db)
gen_symbol<-mapIds(mouse4302.db,keys =as.character(row.names(data_fr)), column = "SYMBOL",keytype = "PROBEID" ) 
T_diff_Th12<-as.data.frame(cbind(gen_symbol,fit$p.value,p.adj))
names(T_diff_Th12)<-c("Gene_symbol","pvalue_Th2_Th1","padj_Th2_Th1")
Th12_sig <- T_diff_Th12[as.numeric(as.character(T_diff_Th12$padj_Th2_Th1))<=0.05,]
write.table(Th12_sig,file = "~/R/Project_3_Lung Fibrosis/comparison_with_TH1_TH2/GSE14308_TH2_VS_TH1_UP_for_TH/GSE14308_family.xml/List of Th1_Th2 without compared with Th17.txt",quote = F)
##p.value <=0.05 is a list of zero elements
```

#Generate the lists including all the information for the Th1/Th2 ratio. The list is going to be used for the comparison of the signature
```{r}
comp_Th1_2 <- cbind(exp_Th1,exp_Th2)
comp_Th1_2$logFC <- comp_Th1_2[,3]- comp_Th1_2[,6]
comp_Th1_2$gene <- mapIds(mouse4302.db,keys=as.character(row.names(comp_Th1_2)),column = "SYMBOL",keytype = "PROBEID" )
comp_Th1_2$direction <- T
comp_Th1_2$direction[comp_Th1_2$logFC<0] <- F
comp_Th1_2 <- comp_Th1_2[,c(8,1:7,9)]
row.names(comp_Th1_2)[4] <- "Th1_mean"
row.names(comp_Th1_2)[7] <- "Th2_mean"
write.table(comp_Th1_2,file =  "~/R/Project_3_Lung Fibrosis/comparison_with_TH1_TH2/GSE14308_TH2_VS_TH1_UP_for_TH/GSE14308_family.xml/Expression_level_of_Th1_versus_Th2.txt",quote = F) 
```

